\documentclass[a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{fullpage}
\frenchspacing
%\pagestyle{empty}
\usepackage[pdftex,bookmarks=false,colorlinks,unicode=true]{hyperref}

\def\mulan{\textsc{MuLan-String}}

\title{MuLan-String library manual}
\author{felix-leg}

\begin{document}

\maketitle
\vfill
    Copyright © 2020  felix-leg. \par
    Permission is granted to copy, distribute and/or modify this document
    under the terms of the GNU Free Documentation License, Version 1.3
    or any later version published by the Free Software Foundation;
    with no Invariant Sections, no Front-Cover Texts, and no Back-Cover Texts.
    A copy of the license is included in the section entitled "GNU
    Free Documentation License".
\pagebreak
\tableofcontents

\pagebreak

\section{Introduction}

\mulan{} (or \textsc{MultiLanguage String}) is an internationalization and localization library made to aid translators whose native language nature forbids simple strings substitution eg. because it uses case system.
The other goal of \mulan{} is to make working with such languages simple for programmers writing programs with i18n in mind. 

The general idea of using \mulan{} is based on \emph{template strings}. The program using the library uses them for building strings from elements. 
The translators formulate template strings which can include rules of how to generate text based on language's case system, genders or pluralization rules.

This manual shows how to use \mulan{} from the view point of programmer and translator. One can read only the part made for one point of view and skip the other. 


\input{ppov}
\input{tpov}
\input{langs}

\pagebreak
\appendix
\input{licence}

\end{document}
